---
title: "assignment1_ques3"
author: "sunny"
date: '2022-02-24'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 3

#### Loading the dataset

```{r}
dogs_data = read.table("dogs.txt",header=TRUE)
Isofluorane = dogs_data$isofluorane
Halothane = dogs_data$halothane
Cyclopropane = dogs_data$cyclopropane

```

### Question 3-A

To check for normality of data distribution across the 3 variables in each record:

Visually, we plot the histogram and qq-plot for each variable to inspect for normality.

Mathematically, we can go through shapiro-wilk normality test to see if the

#### Histograms

```{r}
par(mfrow=c(1,3))
hist(Isofluorane,main="Histogram of Isofluorane data",probability = TRUE,xlab="Conc. in nanongrams per millimeter")
hist(Halothane,main="Histogram of Halothane data", probability = TRUE, xlab="Conc. in nanongrams per millimeter")
#par(mfrow=c(1,1))
hist(Cyclopropane,main="Histogram of Cycloprane data", probability = TRUE, xlab="Conc. in nanongrams per millimeter")
```

Looking visually at histograms, the data doesn't look to be distributed normally across any variable.

#### QQ-Plots

```{r}
par(mfrow=c(1,3))
qqnorm(Isofluorane,main="QQ-Plot of Isofluorane data")
qqnorm(Halothane,main="QQ-Plot of Halothane data")
qqnorm(Cyclopropane,main="QQ-Plot of Cyclopropane data")
```

QQ-Plots reveal a better picture of the normality and looking at it, Halothane and Cyclopropane data look promising candidates for normality.

To confirm our inference, lets go through shapiro-wilk normality test.

#### Shapiro-Wilk Normality Test

```{r}
shapiro.test(Isofluorane)
shapiro.test(Halothane)
shapiro.test(Cyclopropane)
```

Halothane and Cyclopropane data have p-values \> 0.05, which means we can reject the Null Hypothesis H0 that data is not normally distributed. On the other hand, Isofluorane data isn't normally distributed as p-value \<= 0.05.

#### Final Answer

On visual inspection, all 3 data-sets didnt look visually normal using histogram inspection, but halothane and cyclopropane had somewhat straight qq-plots (variation due to low number of observations).

On closer inspection through shapiro-wilk test, Isofluorane proved to be not normally distributed at all but Halothane and Cyclopropane concentrations were indeed normally distributed through the datasets.
